Untwisting the Tetrahelix (v0.5)

### 2017-02-18 doc/TetrahelixUntwist.tex
some observations added

$\theta$ is approximately $131.8103149$ degrees.
The angle $\theta$ is the rotation of a \emph{each} tetrahedra.
That is, a yellow tetrahedron is rotated slightly more than a $1/3$ of a revolution to match the face of the red tetrahedra.
$3 \theta - 2\pi$ is the apparent rotation of $V_3$ relative to $V_0$.

From Robert Gray's site, repeating formula by H.S.M. Coxeter:
\begin{equation}
V(n) =
\left [
  \begin{tabular}{c}
   $ r_{bc} \cos(n \theta) $\\
   $ r_{bc} \sin(n \theta) $\\
   $ n h_{bc}  $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
 $ r_{bc} = \frac{3\sqrt{3}}{10} $\\
 $ h_{bc} = 1/\sqrt{10} $ \\
 $ \theta = \arccos(-2/3) $ \\
  \end{tabular}      
\end{equation}
where $n$ represents each integer numbered node in succession.

@@

This formulation $V(n,c)$ above is valuable, but obscures the essentially fact that the red, yellow, and blue helices distributed
about the central $z$ axis $120\degree$ from each other.
In order to rewrite this expression with an explicit rotation of $2\pi/3$, we expand 
the expression and seek to isolate the term $c2\pi/3$.
\begin{align*}
  \rho_{bc} n + c \theta  &=   \text{\{we aim for 3 in denominator, so we split...\}} \\
    (3 \theta - 2 \pi)n + (c/3)  (\theta /3)  &=   \text{\{we want $2\pi$ in numerator, so add canceling terms...\}} \\
  (3 \theta - 2 \pi)n + (c/ 3) (3 \theta - 2 \pi  + 2 \pi) &=  \text{\{associate...\}} \\
  (3 \theta - 2 \pi)n + (c/ 3) ((3 \theta - 2 \pi)  + 2 \pi) &=  \text{\{distribute...\}} \\  
  (3 \theta - 2 \pi)n + (c / 3) (3 \theta - 2 \pi)  + c 2 \pi /3 &=  \text{\{definition of $\rho_{bc}$...\}} \\
  \rho_{bc} n + (c / 3) \rho_{bc}  + c 2 \pi /3 &=  \text{\{collect like factors...\}} \\  
  \rho_{bc} (n + c/3)  + c 2 \pi /3  \\
\end{align*}
Now the the term on the left is the only one that depends on the scalar $n$. We use this to a create
a new formulation $H_{BCsymmetric}(n,c) = H_{colored}(n,c)$

@@

Although the $\lambda$ parametrization presented above is a sensible one
because it unifies the BC-helix with the equitetrabeam, it is over-specific,
in that it makes a specific choice as to the relationship of the height $h$
to the radius $r$ which is somewhat arbitrary.

@@

\begin{align*}
  \text{one-hop} &= dist(H_{general}(n,c,\rho,d_{\rho}),H_{general}(n,c+1,\rho,d_{\rho}),r_{\rho})  \\
  \text{one-hop} &= dist(H_{general}(0,0,\rho,d_{\rho}),H_{general}(0,1,\rho,d_{\rho},r_{\rho}))  \\  
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + (r_{\rho}\sin(\rho((1/3)\cdot 0 + 0)) - r_{\rho}\sin(\rho((1/3)\cdot 0 + 1)+\frac{2\pi}{3}))^2  +
    (r_{\rho}\cos(\rho((1/3)\cdot0+0)) - r_{\rho}\cos(\rho((1/3)\cdot 0+1) + \frac{2\pi}{3})))^2} \\
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + (0  - r_{\rho}\sin(\rho/3 + \frac{2\pi}{3}))^2  +
    (r_{\rho} - r_{\rho}\cos(\rho/3 + \frac{2\pi}{3}))^2} \\
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}^2\sin^2(\rho/3 + \frac{2\pi}{3})  +
    r_{\rho}^2(1 - \cos(\rho/3 + \frac{2\pi}{3}))^2} \\
  %% Formula above checks at sqrt(8/27), 1, 0, and rho_bc, h_bc, r_bc.
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}^2(\sin^2(\rho/3 + \frac{2\pi}{3})  + (1 - \cos(\rho/3 + \frac{2\pi}{3}))^2)} \\
  d_{\rho}^2 &= 1 - 4 r_{\rho}^2 (\sin( \rho / 2)^2 \\
  %% Formula above checks at sqrt(8/27), 1, 0, and rho_bc, h_bc, r_bc.  
  \text{one-hop}  &= \sqrt{\frac{1}{9}  + r_{\rho}^2(-\frac{4 (\sin^2( \rho / 2))}{9} + \sin^2(\rho/3+ \frac{2\pi}{3})  + (1 - \cos(\rho/3 + \frac{2\pi}{3}))^2)} \\
  %% Formula above checks at sqrt(8/27), 1, 0, and rho_bc, h_bc, r_bc.    
\end{align*}

@@

\begin{align*}
  d_{opt}^2 &= 1 - 4 \frac{2}{\sqrt{\frac{9}{2} \cdot ( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + cos(\rho)+ 8 }}^2 (\sin{ \rho / 2})^2   \\
  d_{opt}^2 &= 1 - 4 \frac{4}{\frac{9}{2} \cdot ( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + cos(\rho)+ 8 } (\sin{ \rho / 2})^2   \\
  d_{opt}^2 &= 1 - \frac{32}{9 ( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + cos(\rho)+ 8 } (\sin{ \rho / 2})^2   \\
  d_{opt}^2 &= 1 - \frac{8 sin^2(\rho) }{ cos(r) + 9( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + 8 }   \\
    d_{opt} &= \sqrt{1 - \frac{8 sin^2(\rho) }{ cos(r) + 9( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + 8 }}    \numberthis  \label{dopt}  \\      
\end{align*}
Therefore, by computing $r_{opt}$ as a function of $\rho$ from this equation, we can construct minimax optimal tetrahelix for an $\rho < \rho_{bc}$.

@@

(Note: This fails the check of computing the BC radius for the BC rail angle.)

@@

\begin{align*}
  \text{rail} &=  1 \\
  \text{one-hop} &= \sqrt{\frac{1}{9} + 3r^2}\\
\text{two-hops}    &= \sqrt{\frac{4}{9} + 3r^2}
\end{align*}
Computing the derivative of two-hops - one-hop:
\begin{align*}
 \frac{\partial \text{two-hops} - \text{one-hop}}{r} &= \frac{\partial \sqrt{4/9 + 3r^2} - \sqrt{1/9 + 3r^2}}{\partial r} \\
  \frac{\partial \text{two-hops} - \text{one-hop}}{r} &= 9r (\frac{1}{\sqrt{27r^2 + 4}} - \frac{1}{\sqrt{ 27r^2 + 1}}) 
\end{align*}
which is negative for all $r> 0$. Thererfore two-hop - one-hop
is minimized when we maximize $r$, up until one-hop = rail.

Note that two-hops increases more and more slowly as $r$ increases. Therefore the minimum maximum will
be obtained with the value of $r$ that when $one-hop = 1$. This can be observed from the
plot of two-hops - one-hops as a function of $r$, or by observing that the derivate of two-hops - one-hop
is negative for $r > 0$. (That is, the value of one-hop increases faster than two-hops.) 
Setting the one-hop distance to $d = 1$ also has the pleasing property of giving us only two lengths. Then
\begin{align*}
   1  &=  \sqrt{\frac{1}{9} + 3r^2} \\
   r  &= \sqrt{\frac{8}{27}} \\
   r &\approx 0.5443
\end{align*}

@@

\section{Adding an Untwisting Parameter}

We observe that by thinking of the straight lines of the
Equitetrabeam as a degenerate helix of zero curvature,
it should be possible to define a smoothly varying continuum between the Boerdijk--Coxeter helix and the Equitetrabeam and every
curvature and torsion between the two. It seems most convenient to define this continuum in terms
of the rail angle $\rho$, where $0 \leq \rho \leq \rho_{bc}$.

Now basic components of the helix, which are the radius $r$, the rate of rotation, and the rate of
axial growth can all be linearly interpolated with a parameter $\lambda$ between their high values (for the BC helix)
and low values (for the equitetrabeam).

\begin{align*}
\phi_{\lambda} &=  \lambda \cdot \rho_{bc}   \\
r_{\lambda}  &=  r_{opt}(\phi) \\
d_{\lambda} &=   d_{opt}(\phi) \\
\end{align*}
to create a formula that generates a continuum of tetrahedral structures:

\begin{equation}
H_{continuum}(n,c,\lambda) =
\left [
  \begin{tabular}{c}
   $ r_{\lambda} \cos(\phi_{\lambda} \kappa + c 2 \pi /3) $\\
   $ r_{\lambda}  \sin(\phi_{\lambda} \kappa + c 2 \pi /3) $\\
   $ d_{\lambda} \kappa $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
    $\kappa = n + c/3$ \\
    $ \phi_{\lambda} =  \lambda \rho_{bc}  $\\
    $\rho_{bc} = (3 \theta - 2 \pi)$ \\
   $ \theta = \arccos(-2/3) $
  \end{tabular}      
\end{equation}

A value of $\lambda = 0$ generates the equitetrabeam, and $\lambda = 1$ generates the Boerdijk--Coxeter helix, and every
value $\lambda \in [-1,1]$ generates an attractive structure which if physically realized is an inherently rigid structure
with member lengths of no greater disparity than $16\%$.
In the sense of structural engineering, it would be a relatively strong space frame. When $\lvert \lambda \rvert > 1$, tetrahelices
are still produced, but as $\lvert \lambda \rvert >> 2$ these tetrahelices probably cease to be interesting to mechanical engineers,
because their ability to withstand axial forces seem to diminish significantly.

@@

 Negative values of $\lambda$ generate clockwise tetrahelices; positive generate counter-clockwise helices.
 (Note: In fact there is something wrong with my math, negative lambda values do produces opposit chirality
 but the lengths are not the same as the positive values.  I need to track this down.)

Furthermore, this formula allows one to design the pitch (in edge length units) as a function of $\lambda$ of the helix (along one rail)
where $\lambda \neq 0$:
\[
p(\lambda) = 2 \pi  \cdot d_{opt}
\]

There is thus an optimal continuum of tetrahelices untwisting between BC and equitetrabeam.
Perhaps surprisingly, this untwisting is accomplished only by changing the length of the two-hop member.

@@

>>>>

$\theta$ is approximately $131.81$ degrees.
The angle $\theta$ is the rotation of a \emph{each} tetrahedra.
That is, a yellow tetrahedron is rotated slightly more than a $1/3$ of a revolution to match the face of the red tetrahedra.
$3 \theta - 2\pi$ is the apparent rotation of $V_3$ relative to $V_0$.

From Robert Gray's site, repeating formula by H.S.M. Coxeter:
\begin{equation}
  \label{graycoxeter}
V(n) =
\left [
  \begin{tabular}{c}
   $ r_{bc} \cos(n \theta) $\\
   $ r_{bc} \sin(n \theta) $\\
   $ n h_{bc}  $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
 $ r_{bc} = \frac{3\sqrt{3}}{10} $\\
 $ h_{bc} = 1/\sqrt{10} $ \\
 $ \theta = \arccos(-2/3) $ \\
  \end{tabular}      
\end{equation}
where $n$ represents each integer numbered node in succession.

@@

This formulation $V(n,c)$ above is valuable, but obscures the essentially fact that the red, yellow, and blue helices distributed
about the central $z$ axis $120\degree$ from each other.
In order to rewrite this expression with an explicit rotation of $2\pi/3$, we expand 
the expression and seek to isolate the term $c2\pi/3$.
\begin{align*}
  \rho_{bc} n + c \theta  &=   \text{\{we aim for 3 in denominator, so we split...\}} \\
    (3 \theta - 2 \pi)n + (c/3)  (\theta /3)  &=   \text{\{we want $2\pi$ in numerator, so add canceling terms...\}} \\
  (3 \theta - 2 \pi)n + (c/ 3) (3 \theta - 2 \pi  + 2 \pi) &=  \text{\{associate...\}} \\
  (3 \theta - 2 \pi)n + (c/ 3) ((3 \theta - 2 \pi)  + 2 \pi) &=  \text{\{distribute...\}} \\  
  (3 \theta - 2 \pi)n + (c / 3) (3 \theta - 2 \pi)  + c 2 \pi /3 &=  \text{\{definition of $\rho_{bc}$...\}} \\
  \rho_{bc} n + (c / 3) \rho_{bc}  + c 2 \pi /3 &=  \text{\{collect like factors...\}} \\  
  \rho_{bc} (n + c/3)  + c 2 \pi /3  \\
\end{align*}
Now the the term on the left is the only one that depends on the scalar $n$. We use this to a create
a new formulation $H_{BCsymmetric}(n,c) = H_{BCcolored}(n,c)$

@@

@@

\begin{align*}
  \text{one-hop} &= dist(H_{general}(n,c,\rho,d_{\rho}),H_{general}(n,c+1,\rho,d_{\rho}),r_{\rho})  \\
  \text{one-hop} &= dist(H_{general}(0,0,\rho,d_{\rho}),H_{general}(0,1,\rho,d_{\rho},r_{\rho}))  \\  
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}\sin(0) - r_{\rho}\sin(rho/3)+\frac{2\pi}{3}))^2  +
    (r_{\rho}\cos(0) - r_{\rho}\cos(\rho/3 + \frac{2\pi}{3}))^2} \\
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + (0  - r_{\rho}\sin(\rho/3 + \frac{2\pi}{3}))^2  +
    (r_{\rho} - r_{\rho}\cos(\rho/3 + \frac{2\pi}{3}))^2} \\
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}^2\sin^2(\rho/3 + \frac{2\pi}{3})  +
    r_{\rho}^2(1 - \cos(\rho/3 + \frac{2\pi}{3}))^2} \\
  %% Formula above checks at sqrt(8/27), 1, 0, and rho_bc, h_bc, r_bc.
  \text{one-hop}  &= \sqrt{\frac{d_{\rho}^2}{9} + r_{\rho}^2(\sin^2(\rho/3 + \frac{2\pi}{3})  + (1 - \cos(\rho/3 + \frac{2\pi}{3}))^2)} \\
  d_{\rho}^2 &= 1 - 4 r_{\rho}^2 (\sin( \rho / 2)^2 \\
  %% Formula above checks at sqrt(8/27), 1, 0, and rho_bc, h_bc, r_bc.  
  \text{one-hop}  &= \sqrt{\frac{1}{9}  + r_{\rho}^2(-\frac{4 (\sin^2( \rho / 2))}{9} + \sin^2(\rho/3+ \frac{2\pi}{3})  + (1 - \cos(\rho/3 + \frac{2\pi}{3}))^2)} \\
  %% Formula above checks at sqrt(8/27), 1, 0, and rho_bc, h_bc, r_bc.    
\end{align*}

@@

\begin{align*}
  d_{opt}^2 &= 1 - 4 \frac{2}{\sqrt{\frac{9}{2} \cdot ( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + cos(\rho)+ 8 }}^2 (\sin{ \rho / 2})^2   \\
  d_{opt}^2 &= 1 - 4 \frac{4(\sin{ \rho / 2})^2}{\frac{9}{2} \cdot ( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + cos(\rho)+ 8 }    \\
  d_{opt}^2 &= 1 - \frac{32(\sin{ \rho / 2})^2}{9 ( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + cos(\rho)+ 8 }    \\
  d_{opt}^2 &= 1 - \frac{8 sin^2(\rho / 2) }{ cos(r) + 9( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + 8 }   \\
    d_{opt} &= \sqrt{1 - \frac{8 sin^2(\rho / 2) }{ cos(r) + 9( \sqrt{3} sin(\rho/3) + \cos(\rho/3)) + 8 }}    \numberthis  \label{dopt}  \\      
\end{align*}
NOTE: I may have made a terrible mistake here, I'm unsure if the numerator should have $sin^2(\rho)$ or $sin^2(\rho/2)$.

Therefore, by computing $r_{opt}$ as a function of $\rho$ from this equation, we can construct minimax optimal tetrahelix for an $\rho < \rho_{bc}$.

@@

@@

\begin{align*}
  \text{rail} &=  1 \\
  \text{one-hop} &= \sqrt{\frac{1}{9} + 3r^2}\\
\text{two-hops}    &= \sqrt{\frac{4}{9} + 3r^2}
\end{align*}
Then:
\begin{align*}
   1  &=  \sqrt{\frac{1}{9} + 3r^2} \\
   r  &= \sqrt{\frac{8}{27}} \\
   r &\approx 0.5443
\end{align*}

@@

To the extent that we value tetrabeams (that is, tetrahelicies with a rail angle of $0$,
and therefore zero curvature) as mathematical or engineering objects
we have motivated the development of $H_{general}$ as a transformation of $V(n)$ defined by
Equation \eqref{graycoxeter} from Gray and Coxeter, as it is difficult to see how
the $V(n)$ 
formulation could ever give rise to a continuum producing the tetrabeam,
since setting the angle in that equation to zero can produce only points on a single
line.

\section{An Untwisted Continuum}

We observe that Equations \eqref{eqrhoopt} and \eqref{dopt} compute $r_{opt}$ and $d_{opt}$ which
create an optimal tetrahelix for any value rail angle $\rho$ between $0$, which
gives the Equitetrahelix and
$\rho_{bc} \approx 35.43 \degree$, which gives the BC Helix.

@@

 Negative values of $\rho$ generate clockwise tetrahelices; positive generate counter-clockwise helices.
 (Note: In fact there is something wrong with my math, negative rho values do produces opposit chirality
 but the lengths are not the same as the positive values.  I need to track this down.)

The pitch of the resulting tetrahix 
where $\rho \neq 0$ is:
\[
p(\rho) = \frac{2 \pi  \cdot d_{opt}}{\rho}
\]

As pitch increases smoothly with decreasing rail angle, it is easy to solve for rail angle
to acheive a given pitch, an ability valuable to engineers or architects who wish
to fit a tetrahelix to given dimension.

Perhaps surprisingly, the optimal untwisting is accomplished only by changing the length of
the two-hop member, leaving the one-hop memever and
rail length equivalenet within this continuum. However, it should be noted that
an engineer or architect may also use $H_{general}$ in the same way, and that minimax
length optimality is a mathematic notion only loosely related to the beauty and
utility of physical structures.

Before deriving Equation \eqref{eqrhoopt}, we created a continuum by using a linear
interpolation between the optimal radius for the Equitetrabeam and the BC Helix. This minimax
optimum of this simpler approach was at most 1\% worse than the optimum computed by
\eqref{eqrhoopt}.

@@

Clean up d_{opt}.

Produce mini table of pitch vs. \rho?

Improve clarity and precision of proofs.

Solve the chirality problem.
